Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-06 latex/paper.tex
slight progress on alg section

\begin{itemize}
	\item subroutine for splitting node is rank-2 NMF
	\item using BCD, subproblem is set of independent NNLS problems with variable vector of length 2
	\item 4 possibilities
	\begin{itemize}
		\item all-zero option never optimum if input is nonnegative
		\item if all-pos option is optimum, it is solution to unconstrained problem
		\item can choose between one-zero options using line 6 of algorithm
		\item describe \cref{alg:r2nmf}
	\end{itemize}
\end{itemize}

@@

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two single-positive-variable solutions.

\Cref{alg:r2nmf} implements this strategy for all rows of $\M{H}$ simultaneously.
It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then determines between the two alternate possibilities as necessary.
\GB{To avoid two passes through the data, we could combine the solve and the update if necessary.  It might also help to pre-compute the inverse of $\M{G}$...  Also Lawton, please check the details of the pseudocode!}

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nmf}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Hierarchical Clustering \cite{KP13}}
\label{alg:hiernmf2}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
		\State \textsc{Split}$({\cal R})$
		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
		\While{size$({\cal Q}) < k$}
			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
		\EndWhile
	\EndFunction
	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
\end{algorithmic}
\end{algorithm}

>>>>

@@

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

\Cref{alg:r2nnls} implements this strategy for all rows of $\M{H}$ simultaneously.
It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then chooses between the two alternate possibilities as necessary.
We note that each row of $\M{H}$ is independent, and therefore this algorithm is easily parallelized.
Solving for $\M{W}$ can be done using inputs $\M{C}=\M{A}\M{H}$ and $\M{G}=\M{H}^\Tra\M{H}$.
\GB{To avoid two passes through the data, we could combine the solve and the update if necessary.  It might also help to pre-compute the inverse of $\M{G}$...  Also Lawton, please check the details of the pseudocode!}

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nnls}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.

@@

\begin{itemize}
	\item summarize approach: expand binary tree until number of leaves hits desired target
	\item choosing leaf node to split is done using priority queue based on that leaf's score
	\begin{itemize}
		\item score can be defined in multiple ways...
	\end{itemize}
	\item explain frontier nodes
	\item explain \Cref{alg:hiernmf2} notation, fields of nodes, etc.
\end{itemize}

\begin{algorithm}
\caption{Hierarchical Clustering \cite{KP13}}
\label{alg:hiernmf2}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaf clusters}
	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
		\State \textsc{Split}$({\cal R})$
		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
		\While{size$({\cal Q}) < k$}
			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
		\EndWhile
	\EndFunction
	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
\end{algorithmic}
\end{algorithm}
